---
title: "STAT 530 Homework 4"
output: html_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

## Instructions

Conduct the following analyses using the `golub.csv` file, available on Canvas. The data are from [https://www.openintro.org/data/index.php?data=golub](https://www.openintro.org/data/index.php?data=golub) and the website describes the rows and columns.

Notice that the working directory in VSCode by default is set to the project root directory. Please set up the correct path to the `golub.csv` file. Run `getwd()` to check the current working directory. Also, you probably need to change the file path when you export the Rmd file.

Please install the `umap` package before running the code. Run the following command in the R console:

```{r}
install.packages("umap")
library(umap)
```

Export your solutions to HTML and submit the HTML file on Canvas.

## Problem 1 (2 points)

Using UMAP, construct two latent numeric features using the 7,129 gene expression values for each patient, then visualize the patients using these two features (1 point), coloring each patient by cancer type (1 point). You can use the default parameters when you run the UMAP algorithm.

\textbf{Solution}.

```{r}
# Load necessary libraries
library(umap)
library(ggplot2)

# Load the dataset
golub <- read.csv("golub.csv", header = TRUE)

# Extract gene expression data (columns 7 onward)
gene_expr <- golub[, 7:ncol(golub)]

# Run UMAP (default parameters)
set.seed(42)  # for reproducibility
umap_res <- umap(gene_expr)

# Create a data frame with UMAP coordinates and cancer type
umap_df <- data.frame(UMAP1 = umap_res$layout[, 1],
                      UMAP2 = umap_res$layout[, 2],
                      Cancer = golub$cancer)

# Plot the UMAP results, coloring by cancer type
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Cancer)) +
  geom_point(size = 2, alpha = 0.8) +
  ggtitle("UMAP Visualization of Golub Data by Cancer Type") +
  theme_minimal()

```

## Problem 2 (3 points)

Calculate the first 5 principal components of the gene expression data, after scaling the data to have mean 0 and variance 1 (1 point). Based on these principal components, use the k-means clustering algorithm to group the patients into three clusters (1 point). Visualize the patients using their UMAP coordinates but this time color them using the k-means cluster assignments (1 point).

\textbf{Solution}.

```{r}
# Scale data (mean 0, variance 1)
scaled_expr <- scale(gene_expr)

# Perform PCA
pca_res <- prcomp(scaled_expr, center = FALSE, scale. = FALSE)  # Already scaled

# Extract the first 5 principal components
X_pca <- pca_res$x[, 1:5]

# Check dimensions (should be 72 rows, 5 columns)
dim(X_pca)

set.seed(42)  # for reproducibility
km <- kmeans(X_pca, centers = 3, nstart = 25)
clusters <- km$cluster

library(umap)
library(ggplot2)

# Run UMAP on the same scaled expression data (or you could use X_pca, but typically
# we do UMAP on the full scaled gene data)
umap_res <- umap(scaled_expr)

# Build a data frame for plotting
umap_df <- data.frame(
  UMAP1 = umap_res$layout[, 1],
  UMAP2 = umap_res$layout[, 2],
  Cluster = factor(clusters)
)

# Plot the UMAP results, coloring by cluster
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  ggtitle("UMAP Visualization Colored by k-Means Clusters") +
  theme_minimal()

```

## Problem 3 (4 points)

Identify genes that are differentially expressed between subjects with AML and those without, adjusting for possible confounding by whether the patient sample came from bone marrow or peripheral blood (1 point). Define a gene to be differential expressed if it has a false discovery rate of at most 0.05 (1 point). How many genes are differentially expressed (1 point)? What is the gene symbol of the most differentially expressed gene, by p-value (1 point)?

\textbf{Solution}.

```{r}

# Create AML indicator and factor for BM/PB
golub$aml_flag <- as.integer(golub$cancer == "aml")
golub$BM.PB <- factor(golub$BM.PB)

# Identify gene columns (columns 7 onward)
genes <- colnames(golub)[7:ncol(golub)]

# Initialize a vector to store p-values
pvals <- numeric(length(genes))
names(pvals) <- genes

# Loop through each gene
for (g in genes) {
  # Extract gene expression
  y <- golub[[g]]
  
  # Skip if zero variance
  if (sd(y) == 0) {
    pvals[g] <- NA
    next
  }
  
  # Build a small data frame for the model
  model_df <- data.frame(
    y = y,
    aml_flag = golub$aml_flag,
    BM.PB = golub$BM.PB
  )
  
  # Safely fit the linear model (catch errors)
  fit <- tryCatch(
    lm(y ~ aml_flag + BM.PB, data = model_df),
    error = function(e) NULL
  )
  
  # If the model was successfully fit
  if (!is.null(fit)) {
    coefs <- summary(fit)$coefficients
    
    # Ensure "aml_flag" row exists in the coefficients table
    if ("aml_flag" %in% rownames(coefs)) {
      pvals[g] <- coefs["aml_flag", "Pr(>|t|)"]
    } else {
      pvals[g] <- NA
    }
  } else {
    pvals[g] <- NA
  }
}

# 1) We've identified p-values for AML effect while adjusting for BM vs. PB.

# 2) FDR adjustment
adj_pvals <- p.adjust(pvals, method = "fdr")

# Mark genes as DE if FDR <= 0.05
de_genes <- names(adj_pvals)[adj_pvals <= 0.05]

# 3) Number of differentially expressed genes
num_de_genes <- length(de_genes)
cat("Number of differentially expressed genes (FDR ≤ 0.05):", num_de_genes, "\n")

# 4) Gene with the smallest (raw) p-value
most_de_gene <- names(which.min(pvals))
cat("Most differentially expressed gene by p-value:", most_de_gene, "\n")

```

